---
title: '`r targets::tar_read(project)`'
subtitle: "Genome-Wide Association Studies"
author: '`r targets::tar_read(author)`'
institute: "_Inserm U1283 / CNRS UMR8199 / Institut Pasteur de Lille / Université de Lille_"
date: '`r format(Sys.time(), "%A, the %d of %B, %Y")`'
output:
  xaringan::moon_reader:
    self_contained: true
    mathjax: null
    css: [assets/umr1283_8199.css, https://cdn.jsdelivr.net/npm/katex@0.13.11/dist/katex.min.css]
    includes:
      in_header: assets/_scripts.html
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---

class: part-slide

```{r setup, include = FALSE}
options("width" = 110)
options(htmltools.dir.version = FALSE)


### Load Packages ==================================================================================
suppressPackageStartupMessages({
  library(here)
  library(knitr)

  library(data.table)

  # library(ragg)
  library(svglite)
  library(showtext)

  library(ggplot2)
  library(ggtext)
  library(patchwork)
  library(scales)

  library(gt)

  library(targets)
  library(katex)

  library(bacon)
  library(utils)
})
```

```{r setup-knitr, include = FALSE}
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval = TRUE, # Default: TRUE
  include = TRUE, # Default: TRUE
  echo = FALSE, # Default: TRUE
  width = getOption("width"),
  comment = "#>",
  fig.align = "center",
  fig.width = 11.5, # Default: 7
  fig.height = 5.75,  # Default: 7
  dpi = 150, # Default: 72
  # dev = "ragg_png"
  dev = "svg"
)
```

```{r setup-ggplot2, code = source(here("scripts/slides/assets/setup-ggplot2.R")), include = FALSE}
```

```{r setup-gt, code = source(here("scripts/slides/assets/setup-gt.R")), include = FALSE}
```

```{r functions, include = FALSE}
eq <- function(tex, d = FALSE, ...) katex::katex_html(tex = tex, displayMode = d, ..., output = "html")

source(here("scripts/tar-utils/tar-pval_trans.R"))

col_numeric_trans <- function(
  palette, domain, na.color = "#808080", alpha = FALSE, reverse = FALSE, trans = NULL
) {
  rng <- NULL
  if (length(domain) > 0) {
    rng <- range(domain, na.rm = TRUE)
    if (!all(is.finite(rng))) {
      stop("Wasn't able to determine range of domain")
    }
  }
  pf <- scales:::safePaletteFunc(palette, na.color, alpha)
  scales:::withColorAttr("numeric", list(na.color = na.color), function(x) {
    x <- trans(x)
    if (length(x) == 0 || all(is.na(x))) {
      return(pf(x))
    }
    if (is.null(rng)) {
      rng <- range(x, na.rm = TRUE)
    }
    rescaled <- scales::rescale(x, from = rng)
    if (any(rescaled < 0 | rescaled > 1, na.rm = TRUE)) {
      warning("Some values were outside the color scale and will be treated as NA", call. = FALSE)
    }
    if (reverse) {
      rescaled <- 1 - rescaled
    }
    pf(rescaled)
  })
}
```

```{r tar-read, include = FALSE}
```

# Methods

---

# Sofware & Parameters

* Variants minimum allele frequency:
    `5.00 %` (using `bcftools view --min-af 0.05`)

* Variants minimum imputation quality `INFO`:
    `0.80` (using `bcftools view --exclude 'INFO/INFO < 0.8'`)

* Minimum/Maximum alleles count per variants:
    `2` (using `bcftools view --min-alleles 2 --max-alleles 2 --types snps`)

* RS ID, HGNC Symbol and Ensembl gene ID:
    * Species: ` `r tar_read(ensembl_species)` `
    * Genome build: ` `r tar_read(genome_assembly)` `
    * Ensembl version: ` `r tar_read(ensembl_version)` `

* Linear/logistic/Firth-regression based association analysis using PLINK:
  * `--vcf file.vcf dosage=DS`, for imputed genotypes dosage.
  * `--mach-r2-filter`, filter out variants with `INFO < 0.1` and `INFO > 2.0`.
  * `--glm`, linear/logistic/Firth-regression based association analysis.
  * `--covar-variance-standardize`, linearly transform numeric covariates to mean-zero, variance one.

```{r software-version, include = FALSE}
suppressWarnings({
  plink_ver <- sub("^v", "", strsplit(system(paste(tar_read(plink2), "--version"), intern = TRUE)[1], split = " ")[[1]][2])
  bcftools_ver <- strsplit(system(paste(tar_read(bcftools), "--version"), intern = TRUE)[1], split = " ")[[1]][2]
})
```

* Software:
    * R ` `r paste0(R.version$major, ".", R.version$minor)` ` ([www.r-project.org](https://www.r-project.org/))
    * PLINK ` `r plink_ver` ` (Chang et al., 2015; doi:[10.1186/s13742-015-0047-8](https://doi.org/10.1186/s13742-015-0047-8))
    * BCFtools ` `r bcftools_ver` ` ([www.htslib.org](http://www.htslib.org/))

---

# Models

```{r models-data, include = FALSE}
df_models <- setDT(tar_read(gwas_models))[
  j = eq := {
    fmt_covariates <- unlist(strsplit(x = covariates, split = " \\+ "))
    if (any(grepl("^PC[0-9]+$", fmt_covariates))) {
      fmt_pcs <- sprintf("PC_%s", paste(
        range(
          as.numeric(sub("PC", "", grep("^PC[0-9]+$", fmt_covariates, value = TRUE)))
        ),
        collapse = "-"
      ))
      sprintf(
        "\\bold{G_{SNV}} = \\bold{%s} + %s",
        raw_trait,
        paste(
          sub(
            "^(.*_)([^_]+)$", "\\1{\\2}",
            c(fmt_covariates[-grep("^PC[0-9]+$", fmt_covariates)], fmt_pcs)
          ),
          collapse = " + "
        )
      )
    } else {
      sprintf(
        "\\bold{G_{SNV}} = \\bold{%s} + %s",
        raw_trait,
        paste(
          sub(
            "^(.*_)([^_]+)$", "\\1{\\2}",
            fmt_covariates
          ),
          collapse = " + "
        )
      )
    }
  },
  by = "tar_group"
]

pp_models <- df_models[
  j = list(
    pp = list(
      ggplot(
        data = rbindlist(lapply(
          X = sprintf("%s (%s)", pretty_trait, tar_group),
          FUN = function(imodel) {
            fread(
              file = unlist(tar_read(
                name = gwas_results_file,
                branches = as.numeric(sub(".*\\((.*)\\)", "\\1", imodel))
              ))
            )[
              order(P)
            ][
              j = `:=`(
                "exppval" = (seq_len(.N) - 0.5) / .N,
                "labels" = paste0(
                  "&lambda;<sub>gc</sub> = ",
                  format(median(qnorm(P / 2)^2, na.rm = TRUE) / qchisq(0.5, df = 1), digits = 3, nsmall = 3)
                ),
                "model" = imodel
              )
            ]
          }
        ))
      ) +
        aes(
          x = .data[["exppval"]],
          y = .data[["P"]],
          colour = sprintf("%s: %s", .data[["model"]], .data[["labels"]]),
          shape = sprintf("%s: %s", .data[["model"]], .data[["labels"]])
        ) +
        geom_abline(intercept = 0, slope = 1, colour = "black", linetype = 2) +
        geom_point(size = 0.60) +
        scale_x_continuous(
          trans =  pval_trans(md = TRUE),
          expand = expansion(c(0, 0.2)),
          limits = c(1, NA)
        ) +
        scale_y_continuous(
          trans = pval_trans(md = TRUE),
          expand = expansion(c(0, 0.2)),
          limits = c(1, NA)
        ) +
        scale_colour_viridis_d(begin = 0.15, end = 0.85) +
        scale_shape_manual(values = rep(1:5, 3)) +
        guides(shape = guide_legend(override.aes = list(size = 2))) +
        labs(
          x = "Expected P-value",
          y = "Observed P-value",
          colour = NULL,
          shape = NULL
        ) +
        theme(
          plot.title.position = "plot",
          plot.caption.position = "plot",
          plot.title = element_markdown(),
          plot.subtitle = element_markdown(face = "italic"),
          axis.text.x = element_markdown(),
          axis.text.y = element_markdown(),
          legend.position = c(0.99, 0.01),
          legend.justification = c("right", "bottom"),
          legend.box.just = "right",
          legend.text = element_markdown(),
          legend.margin = margin(1.5, 1.5, 1.5, 1.5),
          legend.spacing.x = unit(1.5, "pt"),
          legend.spacing.y = unit(1.5, "pt")
        )
    )
  ),
  by = "pretty_trait"
]

bc_models <- df_models[
  j = list(
    bc = list(
      rbindlist(lapply(
        X = sprintf("%s (%s)", pretty_trait, tar_group),
        FUN = function(imodel) {
          bc <- bacon(fread(
            file = unlist(tar_read(
              name = gwas_results_file,
              branches = as.numeric(sub(".*\\((.*)\\)", "\\1", imodel))
            )),
            select = "Z_STAT"
          ))
          data.table(
            "Model" = imodel,
            "Inflation" = inflation(bc),
            "Bias" = bias(bc)
          )
        }
      ))[
        j = "Rank" := rank(rowMeans(cbind(abs(1 - Inflation), abs(Bias))))
      ]
    )
  ),
  by = "pretty_trait"
]
```

```{r print-models-eq, results = "asis"}
for (itrait in unique(df_models[["pretty_trait"]])) {
  cat("*", itrait, "\n")
  for (igroup in unique(df_models[pretty_trait %in% itrait, tar_group])) {
    cat(sprintf("    1. %s\n", eq(
      tex = df_models[df_models[["tar_group"]] %in% igroup, ][["eq"]],
      d = FALSE
    )))
  }
  cat("\n\n")
}
```

```{r print-models-pp, results = "asis", dev = "ragg_png", fig.height = 10, dpi = 120}
for (itrait in unique(df_models[["pretty_trait"]])) {
  cat(sprintf("---\n\n# Models: %s\n\n", itrait))
  cat(".pull-left[\n\n")
  cat(sprintf(
    "Estimates were obtained using the bioconductor R package <code>bacon</code> v%s (doi:%s).\n\n",
    as.character(packageVersion("bacon")),
    '<a href="https://doi.org/10.18129/B9.bioc.bacon">10.18129/B9.bioc.bacon</a>'
  ))
  cat("\n\n")
  print(
    gt(
      data = bc_models[pretty_trait %in% itrait, bc][[1]],
      rowname_col = "Model"
    ) %>%
      umr() %>%
      tab_options(
        column_labels.background.color = "#2a9bd3",
        stub.background.color = "#2a9bd3"
      ) %>%
      # fmt_scientific(columns = "Minimum Corrected P-value", decimals = 2) %>%
      fmt_number(columns = c("Inflation", "Bias"), n_sigfig = 3) %>%
      data_color(
        columns = "Inflation",
        colors = col_numeric_trans(palette = "plasma", domain = NULL, trans = function(x) abs(1 - x))
      ) %>%
      data_color(
        columns = "Bias",
        colors = col_numeric_trans(palette = "plasma", domain = NULL, reverse = TRUE, trans = abs)
      ) %>%
      cols_align("center")
  )
  cat("\n\n]\n\n.pull-right[\n\n")
  print(pp_models[pretty_trait %in% itrait, pp][[1]])
  cat("\n\n]\n\n")
}
rm(pp_models, bc_models)
```

---
class: part-slide

# Results

```{r print-models-results, results = "asis", dev = "ragg_png", fig.height = 5.50}
for (igroup in df_models[["tar_group"]]) {
  cat("\n\n---\n\n")
  cat(sprintf(
    "# Results: `%s (%s)` - Probability-Probability Plot\n",
    df_models[tar_group %in% igroup, pretty_trait], igroup
  ))
  ccat(eq(tex = sprintf("\\footnotesize %s", df_models[tar_group %in% igroup, eq]), d = TRUE))
  cat("\n")
  print(
    tar_read(gwas_results_pp, branches = igroup)[[1]] +
      theme(
        plot.title = element_blank(),
        plot.subtitle = element_blank()
      )
  )
  cat("\n")

  cat("\n\n---\n\n")
  cat(sprintf(
    "# Results: `%s (%s)` - Manhattan Plot\n",
    df_models[tar_group %in% igroup, pretty_trait], igroup
  ))
  ccat(eq(tex = sprintf("\\footnotesize %s", df_models[tar_group %in% igroup, eq]), d = TRUE))
  cat("\n")
  print(
    tar_read(gwas_results_manhattan, branches = igroup)[[1]] +
      theme(
        plot.title = element_blank(),
        plot.subtitle = element_blank()
      )
  )
  cat("\n")

  cat("\n\n---\n\n")
  cat(sprintf(
    "# Results: `%s (%s)` - Volcano Plot\n",
    df_models[tar_group %in% igroup, pretty_trait], igroup
  ))
cat(eq(tex = sprintf("\\footnotesize %s", df_models[tar_group %in% igroup, eq]), d = TRUE))
  cat("\n")
  print(
    tar_read(gwas_results_volcano, branches = igroup)[[1]] +
      theme(
        plot.title = element_blank(),
        plot.subtitle = element_blank()
      )
  )
  cat("\n")

  cat("\n\n---\n\n")
  cat(sprintf(
    "# Results: `%s (%s)` - Top 10 Loci\n",
    df_models[tar_group %in% igroup, pretty_trait], igroup
  ))
  ccat(eq(tex = sprintf("\\footnotesize %s", df_models[tar_group %in% igroup, eq]), d = TRUE))
  cat("\n")
  dt <- fread(
    file = tar_read(gwas_results_file, branches = igroup)[[1]]
  )[
    Symbol != "" & !is.na(Symbol) & P < 0.05
  ]
  effect_name <- grep("^(OR|BETA)$", names(dt), value = TRUE)
  dt <- setnames(x = dt, old = effect_name, new = "EFFECT")[
    j = list(
      SNPs = .N,
      RSID = gsub(",", ";", ID[P == min(P)]),
      EFFECT = EFFECT[P == min(P)],
      P = min(P),
      REF = REF[P == min(P)],
      ALT = ALT[P == min(P)],
      AF = AF[P == min(P)],
      RPAF = RefPanelAF[P == min(P)],
      AC = AC[P == min(P)],
      N = AN[P == min(P)] / 2,
      INFO = INFO[P == min(P)]
    ),
    by = list("Gene" = gsub(",", "/", Symbol))
  ][
    order(P),
    .SD[1:10]
  ]
  print(
    gt(
      data = dt,
      rowname_col = "Gene"
    ) %>%
      umr() %>%
      tab_options(
        column_labels.background.color = "#2a9bd3",
        stub.background.color = "#2a9bd3"
      ) %>%
      tab_style(
        style = cell_text(style = "italic"),
        locations = cells_stub()
      ) %>%
      fmt_scientific(columns = "P", decimals = 2) %>%
      fmt_percent(columns = c("AF", "RPAF"), decimals = 2, incl_space = TRUE) %>%
      fmt_number(columns = c("EFFECT", "INFO"), n_sigfig = 3) %>%
      fmt_number(columns = c("AC", "N"), decimals = 0) %>%
      fmt(columns = "RSID", fns = function(x) gsub(";", "<br>", x)) %>%
      data_color(
        columns = "SNPs",
        colors = col_numeric(palette = "plasma", domain = NULL)
      ) %>%
      data_color(
        columns = "P",
        colors = col_numeric(palette = "viridis", domain = NULL, reverse = TRUE)
      ) %>%
      data_color(
        columns = "EFFECT",
        colors = col_numeric(palette = "plasma", domain = NULL)
      ) %>%
      cols_align("center") %>%
      cols_label(EFFECT = effect_name) %>%
      tab_source_note(
        source_note = html(
          sprintf(
            paste(
              "<b>SNPs</b>, the number of nominally significant SNPs at the locus (&alpha; = 0.05)",
              "<b>RSID</b>, the RSID of the best SNP at the locus (Ensembl v%s)",
              "<b>%s</b>, the %s estimate from <code>PLINK2&nbsp;--glm</code>",
              "<b>P</b>, the p-value from <code>PLINK2&nbsp;--glm</code>",
              "<b>REF</b>, the reference allele",
              "<b>ALT</b>, the alternate allele",
              "<b>AF</b>, the alternate allele frequency",
              "<b>RPAF</b>, the reference panel alternate allele frequency",
              "<b>AC</b>, the alternate allele count",
              "<b>N</b>, the number of genotype/imputed individuals",
              "<b>INFO</b>, the <code>IMPUTE2</code> information score.",
              sep = "; "
            ),
            tar_read(ensembl_version),
            effect_name,
            c("BETA" = "beta", "OR" = "odds ratio")[effect_name]
          )
        )
      ) %>%
      tab_options(table.font.size = "14px")
  )
  cat("\n")
}
```

---
class: part-slide

.center[
<a href="https://www.good.cnrs.fr/" target="_blank"><img src="https://raw.githubusercontent.com/mcanouil/hex-stickers/main/SVG/umr1283_8199.svg" width = "200px"/></br>
<i style="font-size: 200%">www.good.cnrs.fr</i>
</a>
]
