---
title: '`r targets::tar_read(project)`'
subtitle: "Transcriptome-Wide Association Studies"
author: '`r targets::tar_read(author)`'
institute: "_Inserm U1283 / CNRS UMR8199 / Institut Pasteur de Lille / Université de Lille_"
date: '`r format(Sys.time(), "%A, the %d of %B, %Y")`'
output:
  xaringan::moon_reader:
    self_contained: true
    mathjax: null
    css: [assets/umr1283_8199.css, https://cdn.jsdelivr.net/npm/katex@0.13.11/dist/katex.min.css]
    includes:
      in_header: assets/_scripts.html
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---

class: part-slide

```{r setup, include = FALSE}
options("width" = 110)
options(htmltools.dir.version = FALSE)


### Load Packages ==================================================================================
suppressPackageStartupMessages({
  library(here)
  library(knitr)

  library(data.table)

  library(ragg)
  library(svglite)
  library(showtext)

  library(ggplot2)
  library(ggtext)
  library(patchwork)
  library(scales)

  library(gt)

  library(targets)
  library(katex)

  library(bacon)
  library(utils)
})
```

```{r setup-knitr, include = FALSE}
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval = TRUE, # Default: TRUE
  include = TRUE, # Default: TRUE
  echo = FALSE, # Default: TRUE
  width = getOption("width"),
  comment = "#>",
  fig.align = "center",
  fig.width = 11.5, # Default: 7
  fig.height = 5.75,  # Default: 7
  dpi = 150, # Default: 72
  # dev = "ragg_png"
  dev = "svg"
)
```

```{r setup-ggplot2, file = "assets/setup-ggplot2.R", include = FALSE}
```

```{r setup-gt, file = "assets/setup-gt.R", include = FALSE}
```

```{r functions, include = FALSE}
eq <- function(tex, d = FALSE, ...) katex::katex_html(tex = tex, displayMode = d, ..., output = "html")

source(here("scripts/tar-utils/tar-pval_trans.R"))

col_numeric_trans <- function(
  palette, domain, na.color = "#808080", alpha = FALSE, reverse = FALSE, trans = NULL
) {
  rng <- NULL
  if (length(domain) > 0) {
    rng <- range(domain, na.rm = TRUE)
    if (!all(is.finite(rng))) {
      stop("Wasn't able to determine range of domain")
    }
  }
  pf <- scales:::safePaletteFunc(palette, na.color, alpha)
  scales:::withColorAttr("numeric", list(na.color = na.color), function(x) {
    x <- trans(x)
    if (length(x) == 0 || all(is.na(x))) {
      return(pf(x))
    }
    if (is.null(rng)) {
      rng <- range(x, na.rm = TRUE)
    }
    rescaled <- scales::rescale(x, from = rng)
    if (any(rescaled < 0 | rescaled > 1, na.rm = TRUE)) {
      warning("Some values were outside the color scale and will be treated as NA", call. = FALSE)
    }
    if (reverse) {
      rescaled <- 1 - rescaled
    }
    pf(rescaled)
  })
}
```

```{r tar-read, include = FALSE}
```

# Methods

---

# Sofware & Parameters

<!--
* Tissue: ` `r tar_read(ma_params)[["cell_tissue"]]` `
-->

* Linear regression based association analysis using `DESeq2`:
  * `DESeq2::nbinomWaldTest`, Waldt test.
  * `DESeq2::results`, comparisons extraction.
  * `stats::p.adjust(method = "BH")`, false discovery rate (FDR) by Benjamini-Hochberg.

* Software:
    * R ` `r paste0(R.version$major, ".", R.version$minor)` ` ([www.r-project.org](https://www.r-project.org/))

* R packages:
    * `DESeq2` ` `r as.character(packageVersion("DESeq2"))` `
        - [https://bioconductor.org/packages/DESeq2/](https://bioconductor.org/packages/DESeq2/)
    * `bacon` ` `r as.character(packageVersion("bacon"))` `
        - [https://bioconductor.org/packages/bacon/](https://bioconductor.org/packages/bacon/)

---
class: part-slide

# Principal Component Analysis

```{r pca-plots, results = "asis", dev = "ragg_png", fig.height = 10, dpi = 120}
for (pgg in tar_read(twas_pca_plots, branches = 1)[[1]]) {
  cat("---\n\n# Quality-Control Pre-analysis\n\n")
  print(pgg)
  cat("\n\n")
}
```

```{r models-data, include = FALSE}
df_models <- setDT(tar_read(twas_models))[
  j = pretty_trait := factor(pretty_trait, levels = unique(pretty_trait))
][
  j = eq := {
    fmt_covariates <- unlist(strsplit(x = covariates, split = " \\+ "))
    if (any(grepl("^PC[0-9]+$", fmt_covariates))) {
      fmt_pcs <- sprintf("PC_%s", paste(
        range(
          as.numeric(sub("PC", "", grep("^PC[0-9]+$", fmt_covariates, value = TRUE)))
        ),
        collapse = "-"
      ))
      sprintf(
        "\\bold{M_{value}} = \\bold{%s} + %s",
        all.vars(as.formula(paste0("~", raw_trait))),
        paste(
          sub(
            "^(.*_)([^_]+)$", "\\1{\\2}",
            c(fmt_covariates[-grep("^PC[0-9]+$", fmt_covariates)], fmt_pcs)
          ),
          collapse = " + "
        )
      )
    } else {
      sprintf(
        "\\bold{M_{value}} = \\bold{%s} + %s",
        all.vars(as.formula(paste0("~", raw_trait))),
        paste(
          sub(
            "^(.*_)([^_]+)$", "\\1{\\2}",
            fmt_covariates
          ),
          collapse = " + "
        )
      )
    }
  },
  by = "tar_group"
][order(pretty_trait, tar_group)]

pp_models <- df_models[
  j = list(
    pp = list(
      ggplot(
        data = rbindlist(lapply(
          X = sprintf("%s (%s)", pretty_trait, tar_group),
          FUN = function(imodel) {
            fread(
              file = unlist(tar_read(
                name = twas_results_file,
                branches = as.numeric(sub(".*\\((.*)\\)", "\\1", imodel))
              ))
            )[
              order(pvalue)
            ][
              j = `:=`(
                "exppval" = (1:.N - 0.5) / .N,
                "labels" = paste0(
                  "&lambda;<sub>gc</sub> = ",
                  format(median(qnorm(pvalue / 2)^2, na.rm = TRUE) / qchisq(0.5, df = 1), digits = 3, nsmall = 3)
                ),
                "model" = imodel
              )
            ]
          }
        ))
      ) +
        aes(
          x = .data[["exppval"]],
          y = .data[["pvalue"]],
          colour = sprintf("%s: %s", .data[["model"]], .data[["labels"]]),
          shape = sprintf("%s: %s", .data[["model"]], .data[["labels"]])
        ) +
        geom_abline(intercept = 0, slope = 1, colour = "black", linetype = 2) +
        geom_point(size = 0.60) +
        scale_x_continuous(
          trans = pval_trans(md = TRUE),
          expand = expansion(c(0, 0.2)),
          limits = c(1, NA)
        ) +
        scale_y_continuous(
          trans = pval_trans(md = TRUE),
          expand = expansion(c(0, 0.2)),
          limits = c(1, NA)
        ) +
        scale_colour_viridis_d(begin = 0.15, end = 0.85) +
        scale_shape_manual(values = rep(1:5, 3)) +
        guides(shape = guide_legend(override.aes = list(size = 2))) +
        labs(
          x = "Expected P-value",
          y = "Observed P-value",
          colour = NULL,
          shape = NULL
        ) +
        theme(
          plot.title.position = "plot",
          plot.caption.position = "plot",
          plot.title = element_markdown(),
          plot.subtitle = element_markdown(face = "italic"),
          axis.text.x = element_markdown(),
          axis.text.y = element_markdown(),
          legend.position = c(0.99, 0.01),
          legend.justification = c("right", "bottom"),
          legend.box.just = "right",
          legend.text = element_markdown(),
          legend.margin = margin(1.5, 1.5, 1.5, 1.5),
          legend.spacing.x = unit(1.5, "pt"),
          legend.spacing.y = unit(1.5, "pt")
        )
    )
  ),
  by = "pretty_trait"
]

bc_models <- df_models[
  j = list(
    bc = list(
      rbindlist(lapply(
        X = sprintf("%s (%s)", pretty_trait, tar_group),
        FUN = function(imodel) {
          bc <- bacon(fread(
            file = unlist(tar_read(
              name = twas_results_file,
              branches = as.numeric(sub(".*\\((.*)\\)", "\\1", imodel))
            )),
            select = "stat"
          ))
          data.table(
            "Model" = imodel,
            "Inflation" = inflation(bc),
            "Bias" = bias(bc)
          )
        }
      ))[
        j = "Rank" := rank(rowMeans(cbind(abs(1 - Inflation), abs(Bias))))
      ]
    )
  ),
  by = "pretty_trait"
]
```

---
class: part-slide

# Models

```{r print-models-pp, results = "asis", dev = "ragg_png", fig.height = 10, dpi = 120}
for (itrait in unique(df_models[["pretty_trait"]])) {
  cat(sprintf("\n\n---\n\n#  %s\n\n", itrait))
  for (igroup in unique(df_models[pretty_trait %in% itrait, tar_group])) {
    cat(sprintf("\n1. %s\n", eq(
      tex = df_models[df_models[["tar_group"]] %in% igroup, ][["eq"]],
      d = FALSE
    )))
  }
  cat("\n\n")
  cat(sprintf("---\n\n# %s\n\n", itrait))
  cat(".pull-left[\n\n")
  cat(sprintf(
    "Estimates were obtained using the bioconductor R package <code>bacon</code> v%s (doi:%s).\n\n",
    as.character(packageVersion("bacon")),
    '<a href="https://doi.org/10.18129/B9.bioc.bacon">10.18129/B9.bioc.bacon</a>'
  ))
  cat("\n\n")
  print(
    gt(
      data = bc_models[pretty_trait %in% itrait, bc][[1]],
      rowname_col = "Model"
    ) %>%
      umr() %>%
      tab_options(
        column_labels.background.color = "#2a9bd3",
        stub.background.color = "#2a9bd3"
      ) %>%
      # fmt_scientific(columns = "Minimum Corrected P-value", decimals = 2) %>%
      fmt_number(columns = c("Inflation", "Bias"), n_sigfig = 3) %>%
      data_color(
        columns = "Inflation",
        colors = col_numeric_trans(palette = "plasma", domain = NULL, trans = function(x) abs(1 - x))
      ) %>%
      data_color(
        columns = "Bias",
        colors = col_numeric_trans(palette = "plasma", domain = NULL, reverse = TRUE, trans = abs)
      ) %>%
      cols_align("center")
  )
  cat("\n\n]\n\n.pull-right[\n\n")
  print(pp_models[pretty_trait %in% itrait, pp][[1]])
  cat("\n\n]\n\n")
}
rm(pp_models, bc_models)
```

---
class: part-slide

# Results

```{r print-models-results, results = "asis", dev = "ragg_png", fig.height = 5.50}
for (igroup in df_models[["tar_group"]]) {
  cat("\n\n---\n\n")
  cat(sprintf(
    "# `%s (%s)`<i style='font-size:50%%;'> - Volcano Plot</i>\n",
    df_models[tar_group %in% igroup, pretty_trait], igroup
  ))
  cat(eq(tex = sprintf("\\footnotesize %s", df_models[tar_group %in% igroup, eq]), d = TRUE))
  cat("\n")
  print(
    tar_read(twas_results_volcano, branches = igroup)[[1]] +
      theme(
        plot.title = element_blank(),
        plot.subtitle = element_blank()
      )
  )
  cat("\n")

  cat("\n\n---\n\n")
  cat(sprintf(
    "# `%s (%s)` - Top 10 Loci\n",
    df_models[tar_group %in% igroup, pretty_trait], igroup
  ))
  cat(eq(tex = sprintf("\\footnotesize %s", df_models[tar_group %in% igroup, eq]), d = TRUE))
  cat("\n")
  dt <- fread(
    file = tar_read(twas_results_file, branches = igroup)[[1]]
  )[
    external_gene_name != "" & !is.na(external_gene_name) & pvalue < 0.05
  ][
    j = Symbol := paste(unique(unlist(strsplit(gsub(",", ";", external_gene_name), ";"))), collapse = ";"),
    by = "external_gene_name"
  ][
    order(pvalue),
    .SD[1:10]
  ]
  print(
    gt(
      data = setnames(
        x = dt[
          j = .SD,
          .SDcols = c("Symbol", "baseMean", "log2FoldChange", "lfcSE", "pvalue", "fdr", "n", "ensembl_gene_id")
        ],
        old = toupper
      ),
      rowname_col = toupper("Symbol")
    ) %>%
      umr() %>%
      tab_options(
        column_labels.background.color = "#2a9bd3",
        stub.background.color = "#2a9bd3"
      ) %>%
      tab_style(
        style = cell_text(style = "italic"),
        locations = cells_stub()
      ) %>%
      fmt_scientific(columns = toupper(c("pvalue", "fdr")), decimals = 2) %>%
      fmt_number(columns = toupper(c("baseMean"))) %>%
      fmt_number(columns = toupper(c("log2FoldChange", "lfcSE")), n_sigfig = 3, decimals = 2) %>%
      fmt_number(columns = toupper(c("n")), decimals = 0) %>%
      data_color(
        columns = toupper("pvalue"),
        colors = col_numeric(palette = "viridis", domain = NULL, reverse = TRUE)
      ) %>%
      data_color(
        columns = toupper("log2FoldChange"),
        colors = col_numeric(palette = "plasma", domain = NULL)
      ) %>%
      cols_align("center") %>%
      tab_source_note(
        source_note = html(
          paste(
            "<b>LOG2FOLDCHANGE</b>, the estimate from <code>DESeq2</code>",
            "<b>PVALUE</b>, the p-value from <code>DESeq2</code>",
            "<b>FDR</b>, the false discovery rate from Benjamini-Hochberg method",
            "<b>BASEMEAN</b>, the normalised counts from <code>DESeq2</code>",
            "<b>N</b>, the number of individuals.",
            sep = "; "
          )
        )
      ) %>%
      tab_options(table.font.size = if (max(nchar(dt[["Symbol"]])) > 20) "12px" else "14px")
  )
  cat("\n")
}
```

---
class: part-slide

.center[
<a href="https://www.good.cnrs.fr/" target="_blank"><img src="https://raw.githubusercontent.com/mcanouil/hex-stickers/main/SVG/umr1283_8199.svg" width = "200px"/></br>
<i style="font-size: 200%">www.good.cnrs.fr</i>
</a>
]
